---
title: 'Workshop 2: Using ggplot2 for Communication'
author: "Jude Loughran"
date: "2024-05-15"
output: html_document
---
#Workshop2: Using ggplot2 for Communication
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Install and load tidyverse packages
```{r}
library("tidyverse")
```

##Labels 
Good labels on plots can help communicate key findings and comply with formatting requirements of an outlet (scientific journal). In "ggplot2" labels can be added with the "labs()" function.

Let's start with a title.
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=class)) +
  geom_smooth(se.e=FALSE) +
  labs(title="Fuel efficiency generally decreases with engine size")
```
If you want to add more text you can use other functions:
1)"Subtitle" - This adds more details in a smaller font under the title.
2)"Caption" - This adds text at the bottom right of the plot. 
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=class)) +
  geom_smooth(se=FALSE) +
  labs(
    title="Fuel efficiency generally decreases with engine size",
    subtitle="Two seaters (sports cars) are an exception because of their light weight",
    caption="Data from fueleconomy.gov"
  )
```
You can also use "labs()" to replace axis labels and legend titles.
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class)) +
  geom_smooth(se=FALSE) +
  labs(
    x="Engine displacement(L)",
    y="Highway fuel economy(mpg)",
    colour="Car type"
  )
```
For more information read the "ggplot2" help.

#Annotations 
To add text to the plot itself you would use "geom_text()". This works the same way "geom_point()". This is a good function as it lets you to pull data values out from the data itself and include it in the plot.
```{r}
best_in_class<- mpg %>%
  group_by(class)%>%
  filter(row_number(desc(hwy))== 1)

ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class)) +
  geom_text(aes(label=model),data = best_in_class)
```
To handle the label overlap you can use "nudge()" to move the text a certain amount. 

Labels allow for customisation and actions like highlighting the minimum or maximum values which you can get using base R packages.

##Scales 
Changing the default scales on "ggplot2" can help you customise your plots and help to convey the main message of your resutls.

Normally, "ggplot2" automatically adds scales but when you type
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class))
```
"Ggplot2" adds default scales behind the scenes, but if you want to tweak them you can do by offering values to the scale parameters by adding numbers in the appropriate scale arguments.

Remember to provide limits you need to offer a character vector showing the start and end of the limits (e.g.c(0,12)). Note that "?scale_x_continuous()" has alot of other arguments, so don't forget to explicitly state the argument you're providing these limits for (e.g. limits=c(0,12)).
```{r}
ggplot(mpg, aes(displ,hwy)) +
  geom_point(aes(colour=class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

##Axis title 
You can change the ticks on your axes. Breaks control the position of the ticks and you can use labels as an argument to change the text label associated with ticks.
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point() +
  scale_y_continuous(breaks=seq(15, 40, by=5))
```

What does "seq" do?
Try running the code below.
```{r}
seq(15,40,by=5)
```
The code above outputs a sequence of numbers starting at 15 and ending at 40 increasing incrementally by 5.

Similarly, you can use "labels" set to NULL to supress the labels altogether. This is sometimes an option if you're trying to format plots in a certain way.
```{r}
ggplot(mpg, aes(displ,hwy)) +
  geom_point() +
  scale_x_continuous(labels=NULL) +
  scale_y_continuous(labels=NULL)
```

##Legends and colour schemes 
Sometimes if you want to change the position of your legend so you need to control the overall position of the legend. To do this use the "theme()" function. 

Run each line below separately to see how it changes the plot of "base".
```{r}
base <- ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class))
base+theme(legend.position="left")
```

```{r}
base <- ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class))
base+theme(legend.position="top")
```

```{r}
base <- ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class))
base+theme(legend.position="bottom")
```

And now the default
```{r}
base <- ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class))
base+theme(legend.position="right")
```

You can also use "legend.position="none"" to supress the display of the legend altogether.
```{r}
base <- ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class))
base+theme(legend.position="none")
```
##Replacing a scale
It is very useful to plot transformations of the varaibles. For example as we have seen with the diamond proces, it's easier to see the precise relationship between carat and price of we  log transform them using a "ggplot2" function to scale precisely for this purpose. 
There are 2 types of scales you will likely want to switch out:
1)Continuous 
```{r}
ggplot(diamonds,aes(carat,price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```
2)Colour
From a visualisation point of view, this is the most important chagnes you can make to a base "ggplot"
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=drv))

ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=drv)) +
  scale_colour_brewer(palette="Set1")
```
If there are only a few colours, you can add a redundant shape mapping.This will also help ensure your plot is interpretable in black and white.
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=drv,shape=drv)) +
  scale_colour_brewer(palette="Set1")
```
The ColorBrewer scales are documented online at http://colorbrewer2.org/. There are a range of different colour schemes you can import such as this, so read more about them in the ggplot2 cookbook: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/.

When you have predefined colours you want to use you can set them yourself using "scale_colour_manual()".
```{r}
presidential %>%
  mutate(id =33+row_number()) %>%
  ggplot(aes(start,id,colour=party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values= c(Republican="red",Democratic="blue"))
```
To use another package do this. 
Notes:
1)Tibble is a fake dataset.
2)"geom_hex()" is a new geom
```{r}
install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)

df<-tibble( 
  x=rnorm(10000),
  y=rnorm(10000)
)
ggplot(df,aes(x,y)) +
  geom_hex() + 
  coord_fixed()

ggplot(df,aes(x,y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

##Themes 
Can customize the entire theme of your plot. "Ggplot2" has 8 themes by default. There are more are included in add-on packages like ggthemes. (https://github.com/jrnold/ggthemes
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=class)) +
  geom_smooth(se=FALSE) +
  theme_bw()
```

```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=class)) +
  geom_smooth(se=FALSE) +
  theme_light()
```

```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=class)) +
  geom_smooth(se=FALSE) +
  theme_classic()
```

```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=class)) +
  geom_smooth(se=FALSE) +
  theme_dark()
```
One particularly nice thing you can do is to develop your own themes simply by setting some or all of the arguments to "theme()" yourself. Here’s one I wrote so that my plots are always the same style.
```{r}
 theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial")) 
```
##Saving and exporting your plots
Now that you've developed your plot, you want to save it to disk in some format that allows you to import it into your report or to share it with the world. In general, the best way to make a single file will be to export it using "ggsave()".
```{r}
ggplot(mpg,aes(displ,hwy)) + geom_point()
ggsave("my-plot.pdf")
```
--------------------------------------------------------------------------------

