---
title: 'Workshop 2: Using ggplot2 for Communication'
author: "Jude Loughran"
date: "2024-05-15"
output: html_document
---
#Workshop 2: Using ggplot2 for Communication
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Install and load tidyverse packages
```{r}
library("tidyverse")
```

##Labels 
Good labels on plots can help communicate key findings and comply with formatting requirements of an outlet (scientific journal). In "ggplot2" labels can be added with the "labs()" function.

Let's start with a title.
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=class)) +
  geom_smooth(se.e=FALSE) +
  labs(title="Fuel efficiency generally decreases with engine size")
```
If you want to add more text you can use other functions:
1)"Subtitle" - This adds more details in a smaller font under the title.
2)"Caption" - This adds text at the bottom right of the plot. 
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(color=class)) +
  geom_smooth(se=FALSE) +
  labs(
    title="Fuel efficiency generally decreases with engine size",
    subtitle="Two seaters (sports cars) are an exception because of their light weight",
    caption="Data from fueleconomy.gov"
  )
```
You can also use "labs()" to replace axis labels and legend titles.
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class)) +
  geom_smooth(se=FALSE) +
  labs(
    x="Engine displacement(L)",
    y="Highway fuel economy(mpg)",
    colour="Car type"
  )
```
For more information read the "ggplot2" help.

#Annotations 
To add text to the plot itself you would use "geom_text()". This works the same way "geom_point()". This is a good function as it lets you to pull data values out from the data itself and include it in the plot.
```{r}
best_in_class<- mpg %>%
  group_by(class)%>%
  filter(row_number(desc(hwy))== 1)

ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class)) +
  geom_text(aes(label=model),data = best_in_class)
```
To handle the label overlap you can use "nudge()" to move the text a certain amount. 

Labels allow for customisation and actions like highlighting the minimum or maximum values which you can get using base R packages.

##Scales 
Changing the default scales on "ggplot2" can help you customise your plots and help to convey the main message of your resutls.

Normally, "ggplot2" automatically adds scales but when you type
```{r}
ggplot(mpg,aes(displ,hwy)) +
  geom_point(aes(colour=class))
```
"Ggplot2" adds default scales behind the scenes, but if you want to tweak them you can do by.
----------------------------------------
