---
Title: "Workshop 1:Introduction"
Author: "Jude Loughran"
Output: html_document
Date: "14/05/2024"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Installing and Loading Tidyverse
```{r}
library("tidyverse")
```

#Loading the Data
Obtaining the data in order to get the best grasp of "ggplot2" and to fully 
understand the functionality it's vital to remove data interpretation or
uncertainty so we using the build-in "ggplot2" data set. This package's only 
function is to help understand its functionality without worrying about data 
issues, data wrangling or importing data. The mpg data frame in "ggplot2" has 
234 rows of observations and 11 columned variables. To look at the data frame 
"mpg" is typed in the R console. Variables in this data frame include:
1)"displ": the engine size of the cars in L
2)"hwy": the car's fuel efficiency in miles/gallon ("mpg") 
A car with low efficiency uses more petrol compared to a high efficiency car 
when they travel the same distance. To know more about this data set use 
"?mpg" in console.
```{r}
data(mpg)
```
Variables the "mpg" data frame should include:
1)"displ" -  Which is the engine size in L 
2)"hwy" - Which is the car's fuel efficiency in miles/L
A car with a low fuel efficiency uses more fuel than a car with a high fuel 
efficiency when they travel the same distance.

#Quick Data Checks
```{r}
head(mpg)
glimpse(mpg)
summary(mpg)
```
#Creating my First Plot
In this section I am creating my first ggplot, I can plot my loaded data using "data=mpg". I can put the size of the engine ("displ") on the x axis and the fuel efficiency ("hwy") on the y axis . 
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy))
```
The plot shows a negative relationship between engine size and fuel 
efficiency. In other words, cars with big engines use more fuel. What does this say about fuel efficiency and engine size?

When the cars engine size increases the fuel efficiency decreases by 
approximately double. 

"Ggplot()" makes co-ordinate systems that you can add layers to. The first 
argument of "ggplot()" is the data set used in the graph ("data"). So 
"ggplot(data=mpg") makes an empty graph. The next step is to add layers.
The "geom_point()" function adds layer of points to plot which makes a scatter
plot. Each "geom" function takes a mapping argument which defines how 
variables in data set are mapped to visual properties."Geom' is always paired 
with "aes()" and the "x" and "y" specify which variables to map on these axes.
"Ggplot" looks for the map variables in the data argument which in this instance
is "mpg". 

#Understanding the Grammar of Graphics
#Graphing Template
Below is the graphing template that helps you to see what you need to offer "ggplot2" and how you can offer that information. You can use this to help develop "ggplot" plots. 

ggplot(data=<DATA>) +
  <GEOM_FUNCTIO>(mapping=aes(<MAPPINGS>))
  
Your "ggplot" code should always follow the generic template seen above. You will to ALWAYS:
1)Call "ggplot"
2)Input data
3)Provide a "geom" function with a collection of mappings

SO let's revisit this First, we make an empty ggplot() that simply provides the coordinate system that the remainder of the ggplot2 arguments can be mapped to. Try it.
```{r}
ggplot()
```
This just created an empty window because there is no data argument added.
```{r}
ggplot()
```

#Aesthetic Mapping
In addition to plotting co-ordinates "x" and "y", you can add a third variable
like class, by mapping it to an aesthetic represented by "aes()" which can be 
seen in the template above. Aesthetic can include:
1)Size
2)Shape
3)Colour 

When you’re creating a plot, you essentially need two attributes of a plot: 
1)A geom 
2)Aesthetics.
You can convey information about your data by mapping aesthetics in your plot to
the variables in your dataset. To map an aesthetic to a variable you need to 
associate the name of the aesthetic to the name of the variable INSIDE "aes()"

#Change point colour by class
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy,colour=class))
```
Above you have mapped class to colour but you can also map to the size aesthetic in the same way.

NOTE: Don't plot a variable which is not continuous to a continuous 
aesthetic like size.

#Change point size by class
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy,size=class))
```
Note the warning!!!
WARNING: using size for a discrete variable is not advised.

#Change transparency (alpha) by class
Alpha is what dictates the transparency. If you try a simple number like 
0.3 to this aesthetic it will globally change the transparency to all the  
points. This is a useful tip for dense point clouds where you would want a 
reader to see that there are overlapping data points. 

This next body of text is from the "ggplot2 alpha reference material" online:
Alpha refers to the opacity of a geom. Values of "alpha" range from 0-1, with 
the lower values referencing to more transparent colours. Alpha can also be 
modified by "colour" or "fill" if either aesthetics provide colour values using
an RGB specification like "RRGGBBAA" where "AA" references the transparency 
values.
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy,alpha=class))
```
When would using alpha be appropriate?
Using alpha would be appropriate when:
1)Trying to visualize data 
2)Highlighting data points 
3)Layering data points 
4)Customizing data 

#Change point shape by class
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy,shape=class))
```
You can also set properties manually like offering a number/colour.
 
#Make all points blue
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy),color="blue")
```
There is a range of manual aesthetics that can be set to manually change 
appearance of plot:
1)Name of a colour as a character string 
2)Size of a point in mm
3)Shape of a point as a number 

What happens if try to map aesthetic to something other than a variable name 
like "aes(colour=displ<5)?
When this was tried a ggplot was produced where data points were separated into 
teal and red colours based on whether their fuel efficiency was above or below 5 miles/L. If it was above the data point was teal and if it was below the data point was red. See below.
```{r}
ggplot(data=mpg) +
  geom_point(mapping = aes(x=displ,y=hwy,colour=displ<5))
```

#Troubleshooting 
Below there was a purposeful mistake in the code, there was an extra "+"  on top line that I removed. Below is the fixed code.
```{r}
ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ,y=hwy)) 
```

#Faceting and Panal Plots
It is common practice to break single complex plots into many sub plots (panels). This allows for development of separate plots for a range of reasons. In "ggplot" this is done using "facets". To "facet" a plot using 1 variable, use "facet_wrap()" and "~" which dictates which variable you want to subset data with. 
NOTE: is for discrete variables only.

#Facet_wrap
Split the plot out by car type (class).
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y =hwy)) + 
  facet_wrap(~class,nrow=2)
```
#Facet_grid
If you want to more than one variable use "facet_grid()". Here need two variables and you can use "~" to split them up. A separate facet for each combination of drive-type (e.g. 4WD) * number of cylinders
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy)) + 
  facet_grid(drv~cyl)
```
NOTE: There are no occurrences of 5 cylinder 4WDs OR RWD vehicles!

#Facet_grid by just Rows or Columns
Using "." is useful if you don't want to facet in row or column dimensions. See what happens below with

ggplot(data=mpg) +
  geom_point(mapping=aes(x=displ,y=hwy)) +
  facet_grid(.~cyl).

```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy)) + 
  facet_grid(cyl~.)
```
Or alternatively you can use "facet_grid(.~cyl).
```{r}
ggplot(data = mpg) +   
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(.~cyl)
```

#Exercise 1
Read "?facet_wrap".
```{r}
?facet_wrap
```
What does "nrow" and "ncol" do? 
These control the number of rows and columns in the plot.

What other options control the layout of the individual panels? 
The below chunk of code changes the order in which the panels appear and changes
the levels of the underlying factors.
```{r}
mpg$class2 <- reorder(mpg$class, mpg$displ)
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(vars(class2))
```

This next chunk of code you can allow scales to vary across the panels with 
the "scales" argument. Free scales make it easier to see patterns within each 
panel, but it harder to compare across panels.
```{r}
ggplot(mpg, aes(displ,hwy)) +
  geom_point() +
  facet_wrap(vars(class),scales="free")
```

#Lines
"Ggplot" can use variety of visual objects to show your data. To show data as
points use "geom_point()" which we have already used.
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy))
```

#Geom_smooth
To display the same data as a smooth line fit through the points use geom_smooth(). However, ggplot2 can use a variety of geom objects to represent the data. Here, we might want to use bar plots, line charts, boxplots and so on. Well we can handle this issue in ggplot directly using a different geom to plot the same data. Here, instead of plotting points, we will use a smooth line. 
```{r}
ggplot(data=mpg) + 
  geom_smooth(mapping=aes(x=displ,y=hwy))
```

Why don't we use geom_line() here? What would that look like?
We don't use "geom_line()" here because this package creates line plots and adds a line that goes through connecting dots. See below.
```{r}
ggplot(data=mpg) + 
  geom_line(mapping=aes(x=displ,y=hwy))
```

So let’s recap. A geom is an object that your plot uses to represent the data. 
To change the geom type in your plot, simply change the geom function that you 
add to your plot template. Sometimes you may want to try a few things out, 
in which case you could use comments to help you remember what worked and what 
didn’t.

#Using Comments
Comments are very useful as they allow the user to keep track of the work they
have done or any errors they have fixed. It also allows readers to follow their
coding methodology with more precision and clarity. See below
```{r}
ggplot(data= mpg) + 
  #geom_point(mapping=aes(x=displ,y=hwy)) + #points horrible 
  geom_smooth(mapping=aes(x=displ,y=hwy)) #try smooth line
```
How does R know where to plot the line?
Can use the chunk output
to inform us and also use the help menu.

R knows where to plot the line in accordance to the relationship between the variables specifed in aes()".

#Changing the Linetype
When using "drv" it sorts cars into 3 types based on type of car.
```{r}
ggplot(data=mpg) + 
  geom_smooth(mapping=aes(x=displ,y=hwy,linetype=drv)) 
```
You can also use "lty =" for the same result.
```{r}
ggplot(data=mpg) + 
  geom_smooth(mapping=aes(x=displ,y=hwy,lty=drv)) 
```

#Grouping
Using the "group" argument to show that the data is grouped. You can set the 
"group" aesthetic to a categorical variable to draw multiple objects.
```{r}
ggplot(data=mpg) +
  geom_smooth(mapping=aes(x=displ,y=hwy,group=drv))
```
Look at "ggplot2" cheat sheet for more tips.

#Changing the Line Colour bbased on "drv" Value
```{r}
ggplot(data=mpg) +
  geom_smooth(
    mapping=aes(x=displ,y=hwy,color=drv),
    show.legend=FALSE,
  )
```

#Multiple Geoms
We are increasing complexity by plotting multiple geoms on a single plot.
All that is needed to be done is add them together. This is a nice choice for 
showing the underlaying data and how it relates to the "geom_smooth" line.
We already did this one way.
```{r}
ggplot(data=mpg) + 
  geom_point(mapping=aes(x=displ,y=hwy)) +
  geom_smooth(mapping=aes(x=displ,y=hwy))
```

Duplicating the top row means if wanted to change the "x" variable it would 
need to be changed in other areas. This is not ideal in a programming sense and 
can increase chance of an error so "ggplot" which allows for overarching 
mappings to its argument making them GLOBAL mappings when applied to every "geom". A better way is shown below.
```{r}
ggplot(data = mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_point() + 
  geom_smooth()
```

#Manipulating each Geom Separately
Now using mappings in specific layers to enable the display of different 
aesthetics in different layers.
```{r}
ggplot(data=mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_point(mapping=aes(colour=class)) + 
  geom_smooth()
```

Now we have different colours to indicate the raw data (geom_point) and one 
colour for the overall relationship (geom_smooth).

Specifying further using class=("subcompact")
```{r}
ggplot(data=mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_point(mapping=aes(color=class)) + 
  geom_smooth(data=filter(mpg,class=="subcompact"),se=FALSE)
```
What does the "se = FALSE" do?
"se=false" makes sure the standard error bars are plotted without additional 
standard error bands.

#Exercise 2:
1.What geom would you use to draw a line chart? 
"geom_line"
1.1.A boxplot? 
"geom_boxplot"
1.2. histogram? 
"geom_histrogram
1.3.An area chart?
"geom_area"

2.Run this code in your head and predict that the output will look like. 
Then, run the code in R and check your predictions.
```{r}
ggplot(data=mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_boxplot(mapping=aes(color=class)) 

ggplot(data=mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_histogram(mapping=aes(color=class)) 

ggplot(data=mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_area(mapping=aes(color=class)) 
```

3.Will these two graphs look different? Why/why not?
These 2 graphs will look the same because they codes are defining the same 
parameters, it's just the top code is the more efficient way of coding.
```{r}
ggplot(data=mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_smooth(data=mpg,mapping=aes(x=displ,y = hwy))
```

#Transformations and Statistics
Learning basic transformations and data summaries using a second "ggplot" data.
"Diamonds" has ~54,000 diamonds including:
1)"Price"
2)"Carat"
3)"Colour"
4)"Clarity"

#Plotting statistics 
#Loading Statistics
```{r}
data("diamonds")
glimpse(diamonds)
```
#Bar Plots
This next part straight from textbook:
On x-axis the cut is shown and on the y-axis the count is shown but count is 
not a variable of "diamonds". Where does in come from?
1)Bar charts, histograms, and frequency polygons “bin” data and then plot bin 
counts (the number of points that fall in each bin). 
2)Smoothers fit a model to your data and then plot predictions from the model. 
3)Boxplots compute a robust summary of distribution and then display a specially 
formatted box.
```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut))
```

The algorithm to calculate the new values for a graph is called a "stat" and
Can use "geoms" and "stat" interchangeably. Below I've re-created a previous 
plot using "stat_count()" instead of "geom_bar().
```{r}
ggplot(data=diamonds) + 
  stat_count(mapping=aes(x=cut))
```

#Overriding Defaults
Might want to override a default stat and change the default stat 
(a count or summary) to identity (raw value of a variable). Tribbles is 
essentially a dataframe . Make some new data
```{r}
demo <- tribble(
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo
```
```{r}
ggplot(data=demo) +
  geom_bar(mapping=aes(x=cut,y=freq),stat="identity")
```
Why did we specify now specify a y axis?
On the "y" axis we are plotting a second variable. 

#Proportion rather than the Absolute Frequency
We can also override a default mapping from transformed variables to aesthetics.
EXAMPLE: you could display a bar chart of the PROPORTION of your total diamond 
dataset rather than a count.
```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut,y=stat(prop),group=1))
```
Does anyone get the warning that's in the workbook?
Warning: "stat(prop)"` was deprecated in ggplot2 3.4.0.
If so, can use "geom_bar(mapping = aes(x = cut, y = stage(after_stat = prop), 
group = 1))"
No I did not get a warning.

#Plotting Statistical Details
You can also use the "stat_summary" argument to show more about transformations.
```{r}
ggplot(data=diamonds) + 
  stat_summary(
    mapping=aes(x=cut,y=depth),
    fun.min=min,
    fun.max=max,
    fun=median
  )
```

#Aesthetic Adjustments
Another way to boost the way you can convey information with plots using ggplot2 is to use aesthetics like colour or fill to change aspects of bar colours. We already did this once, but there are multiple options available to you, including changing the fill our outline colours.
```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut,colour=cut))
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut,fill=cut))
```
Does anyone notice anything different in the colour scheme of this plot?
(Hint: It's in the viridis colour palette (colour blind friendly), but why is it
different from the colour palette we used earlier?)
The colours in the viridis palette are darker then the ones we used before.

Check out the difference:
```{r}
ggplot(data = mpg,mapping=aes(x=displ,y=hwy)) + 
  geom_point(mapping=aes(color=class)) 

ggplot(data=diamonds,mapping=aes(x=color,y=depth)) + 
  geom_point(mapping=aes(color=cut)) 
```

#Filling by a Variable
Now try using these aesthetics to colour by another variable like clarity. 
Notice how the stacking is done automatically. This is done behind the scenes
with a position argument.
```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut,fill=clarity))
```

#Altering Transparency (alpha)
Ability to make "position adjustments" is crucial as allows your customise plots
in 4 ways:
1)"Identify" (raw data): Using "position="identity"" able you to place each object EXACTLY where it in the context of the graph. This is vital for chart points (scatter) but does not work for bar plots as too much information. So in this case need to change the bar aesthetic. 
```{r}
ggplot(data=diamonds,mapping=aes(x=cut,fill=clarity)) + 
  geom_bar(alpha=1/5,position="identity")
```

To color the bar outlines with no fill color.
```{r}
ggplot(data=diamonds,mapping=aes(x=cut,colour=clarity)) + 
  geom_bar(fill=NA,position="identity")
```

#Position Adjustments
2)"Fill" (changes heights): Position="fill" works like stacking, but makes each set of stacked bars the same height.
```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut,fill=clarity),position="fill")
```

#Position="dodge" 
3)"Dodge" (doesn't put things on top of each other):"Position=dodge"" places overlapping objects directly beside each other places overlapping objects directly beside one another.
```{r}
ggplot(data=diamonds) + 
  geom_bar(mapping=aes(x=cut,fill=clarity),position="dodge")
```

#Jittering
4)"Jitter": Position = "jitter" adds a small amount of random noise to each point to avoid overplotting when points overlap. This is useful for scatterplots but not barplots.
```{r}
ggplot(data=mpg) + 
  #geom_point(mapping=aes(x=displ,y=hwy),position="jitter")
  geom_point(mapping=aes(x=displ,y=hwy))
```
----------------------------------
#Plot reconstruction 

#First step is to save data into the data folder

#Second step is to import data 
```{r import}
library(readr)
SharkMeta <- read_csv("data/SharkMeta.csv")
View(SharkMeta)
```

#Third step create a basic barplot
```{r}
ggplot(data=SharkMeta, aes(x=SEX)) +
  geom_bar(fill = "pink", color ="purple") +
  labs(title ="Distribution of Shark Sexes",
       x="Sex",
       y="Individual tag count") + #Change "y" label
  theme_minimal() +
```

#Fourth step create a more detailed barplot
```{r shark barplot}
#Define your custom color palette
my_colors <- c("Female" = "pink", "Male" = "lightblue", "Unknown" = "gray")

#Create the bar plot with custom colors
ggplot(data = SharkMeta, aes(x = SEX)) +
  geom_bar(fill = my_colors, color = "purple") +
  scale_x_discrete(labels = c("F" = "Female", "M" = "Male", "U" = "Unknown")) +
  labs(title = "Distribution of Shark Sexes",
       x = "Sex",
       y = "Individual tag count") +
  theme_minimal() +
  theme(axis.text = element_text(size = 14, color = "black", face = "plain"),
        axis.title = element_text(size = 14, color = "black", face = "bold"),
        axis.line = element_line(color = "black", size = 0.5),
        panel.grid.major = element_line(color = "gray", linetype = "dotted"),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "white", color = "black"))
```
